---
title: "10_Intro_plotting_R"
author: "JR"
date: "12/7/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)
library(tidyverse)
source("util/plotting_functions.R")
source("util/_setup.R")

```

One of the best parts of R is the plotting abilities. You are, by analogy designing your own figures with code. You can run many stastical analyses and plot the outputs in the code to make a figure.

Here we will focus on GGPLOT(GGPLOT2). The GG says it all:

"Grammar of Graphics" -- it truly is a very pure and flexible way of plotting data for figures!


Although this is a package it is the best for vast majority of figure making.
There are base R functions that can read excel files etc, but we hopefully will be so clean and tidy we won't be loading excel files into R :) 

There are 8 possible layers of information -- each building on the data layer that you are importing. This looks really confusing but we will slowly build up from the basics (mapping, stats, geom)

You can also think of ggplot2 as an API. There are alternatives such as Vega-lite that are also based in graphical grammar.


Plotting layers of R:

1) Data -- data being plotted from object indexes etc.

2) MAPPING -- mapping data (aes)

3) Statistics -- statical analyses on mapped data

4) scales -- how big small you want things

5) GEOMETRIES -- what type of plots you are making (geom)

6) Facets -- making multiple representations of plots

7) Coordinates -- specifing placement of image aspects

8) THEME -- a set version of colors, background. Mostly so you don't have to add many of the layers above everytime!

The nice thing, to me, is that these layers don't have to come in order. But many of the later layers require data already to be mapped in aes().

Ok let's walk through some simple examples to more difficult figures using these layers!

Layer 1 : DATA (data types matter so good to label object accrodingly)

```{r}
# before we get started let's import that data we are going to     # plot. We will use the peak occurent matrix from our previous
# general analysis of the data previously (10_our_data_ranges).

num_peaks_df <- read_csv('chipseq/results/num_peaks_df.csv')
# There are 22 DBPs in here that don't have values
# for the tf and dbp column -- let's take those
# out since we'll be plotting those columns frequently
num_peaks_df <- num_peaks_df %>% filter(!is.na(tf))

table(num_peaks_df$tf)

# Cool, take a look and we can start plotting these columns !
```




```{r}
# let's make this simple plot first

ggplot(num_peaks_df, aes(x = num_peaks,
                         y = total_peak_length)) +
  geom_point()

# here we are just looking at a x, y point plot (geom_point).
# note there are not quotes on x, and y objects as they are being evaluated as "objects in ggplot"
# note the +, we will see this everytime we are adding a new layer. # Yet within each layer it's standard comma 
# seperated etc.
# '+' adds layers together

             
# There is lots more we could modify here but let's change the    # shape of the points to show off R plotting. We need to do this  # outside the aes() layer. The aes is just for anything for       # that depends on any column of data. 

ggplot(num_peaks_df) +
  geom_point(aes(x = num_peaks, 
                 y = total_peak_length),
             shape = 'square',
             color = 'purple') 

# what other shapes are there and how would you find out ??????

# Now the points are plotted as squares.
  

?geom_point 
# scroll to the Aestheitcs section for all options
# cool let's make the squares more transparent for example

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length,
                 color = tf)) +
  geom_point()

# Now color came in aes() because it depends on data values.

# so much can be done in the Aesethics mapping layer!

# We can also add some color features in the mapping layer of graphical grammar let's take a look
abundant_dbds <- table(num_peaks_df$dbd) > 100
names(which(abundant_dbds))
# we are checking which DBDs have more than 100 enteries or
# we have at least 100 DBPs with a given DBD.
num_peaks_df$dbd == "C2H2 ZF"
# here is the computation we are going to do in the aes layer
# we are going to see above how to take a list of characters
# and turn it into a boolean vector (T/F). Let's go over above
# and see how we can just add to the aes below.
ggplot(num_peaks_df) +
  geom_point(aes(x = num_peaks, 
                 y = total_peak_length,
                 color = dbd == "C2H2 ZF"))
table(num_peaks_df$dbd)
# so we did a bit extra by adding a "function of less-than 1,000 peaks". 
# Thus R will know to add two colors to data with less than 1,000 peaks 
# and another color for those over 1,000 peaks.
# Finally note ggplot gives us a free title and legend!

```


# Histogram. This is great when you are inloadterested how often one variable occurs. So you would only use "x"
# or one data value in this case.

```{r}

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_histogram()

# let's add more bins

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_histogram(bins = 30)

# Let's chage the color of the bars in histogram by number of peaks.

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
  geom_histogram(bins = 30)

# this is hard to see the two distributions.

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_histogram(bins = 30, position = "dodge")

# so now we can see the individual data points seperated!

# let's move on to check out density 
# DENISTY Plots are also very useful to see where the denisty of  # data arises.

# let's take a look at density plots comparing those that are and 
# are not TFs

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_density()

# but again its hard to see let's fix that with shading or alpha

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_density(alpha = 0.1)

# let's change color of the line now.

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf, color = tf)) +
   geom_density(alpha = 0.3)

# so now line and fill are same color -- let's see:

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf, color = tf)) +
   geom_density(alpha = 0.0)

# We can even add layers of plots. For example a 2 dimensional denisty plot of our x, y point plot above.
# Let's check it out

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
  geom_point()+
  geom_density_2d() 

# what would happen if we change the order?

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
   geom_density_2d()+
    geom_point()

# it's hard to see but in this case the contour is on top
# of points and points are below -- subtle but important to       # remember

```

You can do a lot in the geometry layer and add features with + and then new geom call. Let's look at 
# this below


```{r}
# Now let's see how to add other features to the plots by adding an additional geom layer as we did above
# Lets explore 'geom_abline' geometry to add a line 
?geom_abline
# we usually give slope and or intercept

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
  geom_point()+
  geom_abline(slope = 1000, intercept = 0)

# cool this fits -- what does it mean?
# hint average peaks size is 1kb

```

Let's take a quick look at bar plots
```{r}
# Here we will look at more aspects of graphical grammar with bar plots

ggplot(num_peaks_df, aes(x = dbd)) + 
  geom_bar()

  
# If you want the heights of the bars to represent values in the 
# data, use 'stat = identity' and provide that identity in the
# y = statement in aes.

# If the we want to provide the Y-axis for a specific number
# we need to apply stat = identitiy in the geom_bar()

ggplot(num_peaks_df, aes(
      x = dbp,
      y = num_peaks)) +
        geom_bar(stat = "identity")

```

Now we have gone through these layers:

1) Data -- data being plotted from object indexes etc.

2) MAPPING -- mapping data (aes)

3) Statistics -- statical analyses on mapped data

5) Geom

# Now let's look at the scales layer

4) scales -- how big small you want things
# Scales is important for outputting specific factors as colors that are "scaled"
# alos other outputs need to be scaled such as scale values to the "mean" etc.
?scale

```{r}
# Let's use our num_peaks versus peak width to see this
# Todo color by lncRNA and mRNA 
ggplot(num_peaks_df, aes(x =dbp, y =total_peak_length, color = tf)) +
  geom_point()+
  scale_color_brewer(type = 'qual')
# qual is a type scale_colour_brewer that has many color palattes
# We can also set Pallete directly:

ggplot(num_peaks_df, aes(x =dbp, y =total_peak_length, color = tf)) +
  geom_point()+
  scale_color_brewer(palette =  2)

# Try a few numbers.

  

#Let's look of using the scales layer by sclaing the axes

ggplot(num_peaks_df, aes(x = num_peaks, y =total_peak_length, color = tf)) +
  geom_point()+
  scale_color_brewer(palette =  2) +
  scale_x_continuous(breaks = c(1e3, 1e4, 1e5, 1e6)) +
  scale_y_continuous(trans = 'log10')

# TODO explain


# what if we want to set the limits of the axes?

ggplot(num_peaks_df, aes(x = num_peaks, y =total_peak_length, color = tf)) +
  geom_point()+
  scale_color_brewer(palette =  2) +
  xlim(250, 2e3) +
  ylim(1e5, 1e6)

# coool so we zoomed in 



```

Now let's look at 6) FACETS
This will allow us to make multiple plots in the same plot frame or facet
One limitation is that it can only make the same plots from different data


```{r}
#TODO Debug
table(num_peaks_df$dbd)[order(table(num_peaks_df$dbd))]
dbds <- c("C2H2 ZF", "Homeodomain", "bZIP")
ggplot(num_peaks_df %>% filter(dbd %in% dbds),
       # this is cool we can run function on df before 
       # pushing to aes()
       aes(x = num_peaks, y = total_peak_length )) +
  facet_grid(dbd ~ tf) + 
  geom_point()
# compare facet wrap

ggplot(num_peaks_df %>% filter(dbd %in% dbds), 
       aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(dbd ~ tf) + 
  geom_point()

# difference all combos of each factor.

# now let's have them scale themselves using scales = free.
# In facet grid one axis has to be shared, but in facet wrap we 
# can autoscale both axes. We can also tell it the number of 
# rows or columns we want. Let's see below.

# first let's just look at facet_wrap
ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ .) + 
  geom_point()

# here we change the ordering of the figures
ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ ., nrow = 2) + 
  geom_point()

# here we are ordering and allowing each plot 
ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ ., scales = "free", ncol = 1) + 
  geom_point()

# ncol and nrow

```

7) Coordinates
How to represent X and Y coordinates in different coordinate systems.

```{r}

# TODO better coordinates example than polar :)?

ggplot(num_peaks_df, aes(x = dbd)) +
  geom_bar() +
  coord_polar()

# We can play around in here too to tell what is the "theta" or circular dimension

ggplot(num_peaks_df) +
  geom_bar(aes(x = dbd)) +
  coord_polar(theta = 'y')



```

8) Theme
This is a great way to package all the features you want, colors,
backgrounds, grid lines etc. We tend to use paper white as a defualt theme.

We made a defualt theme that we can source at the begining of the
document call _setup.R. Once loaded a "paper white" theme will be available.

Let's look at example from 10_our_data_ranges

```{r}
ggplot(num_peaks_df, aes(x = num_peaks)) + 
  geom_density(alpha = 0.2, color = "#424242", fill = "#424242") +
  theme_paperwhite() +
  xlab(expression("Number of peaks")) +
  ylab(expression("Density")) +
  ggtitle("Promoter binding events",
          subtitle = "mRNA and lncRNA genes") 

# we can see here how titles can easily be made with ggtile which
# is in the "theme" layer of things.

# There are a number of built-in themes in addition to 
# the one we've provided. You can see them by typing theme_ 
# then autocomplete and check a few out! Here is cleavland :

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_density() +
  theme_cleveland()


# Go take a look at _setup.R and see the features of paper white!
```

Class exercise, go back to one of the other figures from 10_our_data_ranges. Try and change some feature or add a regression line etc. Slack us what you get :)

