---
title: "10_Intro_plotting_R"
author: "JR"
date: "12/7/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)
library(tidyverse)
source("util/plotting_functions.R")
source("util/_setup.R")

```

One of the best parts of R is the plotting abilities. You are, by analogy designing your own figures with code. You can run many stastical analyses and plot the outputs in the code to make a figure.

Here we will focus on GGPLOT(GGPLOT2). The GG says it all:

"Grammar of Graphics" -- it truly is a very pure and flexible way of plotting data for figures!


Although this is a package it is the best for vast majority of figure making.
There are base R functions that can read excel files etc, but we hopefully will be so clean and tidy we won't be loading excel files into R :) 

There are 8 possible layers of information -- each building on the data layer that you are importing. This looks really confusing but we will slowly build up from the basics (mapping, stats, geom)

You can also think of ggplot2 as an API. There are alternatives such as Vega-lite that are also based in graphical grammar.


Plotting layers of R:

1) Data -- data being plotted from object indexes etc.

2) MAPPING -- mapping data (aes)

3) Statistics -- statical analyses on mapped data

4) scales -- how big small you want things

5) GEOMETRIES -- what type of plots you are making (geom)

6) Facets -- making multiple representations of plots

7) Coordinates -- specifing placement of image aspects

8) THEME -- a set version of colors, background. Mostly so you don't have to add many of the layers above everytime!

The nice thing, to me, is that these layers don't have to come in order. But many of the later layers require data already to be mapped in aes().

Ok let's walk through some simple examples to more difficult figures using these layers!

Layer 1 : DATA (data types matter so good to label object accrodingly)

```{r}
# before we get started let's import that data we are going to     
# plot. We will use the peak occurent data frame from our previous
# general analysis of the data previously (10_our_data_ranges).


# Here we want a "categorical variable" for plotting colors etc. So we 
# are first going to filter all DBPs that are not marked as DNA binding
# proteins or the 'tf' column is yes.

num_peaks_df <- read_csv('analysis/01_global_peak_properties/results/num_peaks_df.csv')

peak_occurence_df <-

#num_peaks_df_2020 <- read_csv('../../data/2020_peak_occurence_df.csv')

#num_peaks_df <- num_peaks_df %>% filter(!is.na(tf))

# ?is.na
# Note the ! almost always will mean NOT


```


# Cool, take a look and we can start plotting these columns !

```{r}
# let's make this simple plot first

ggplot(num_peaks_df, aes(x = num_peaks,
                         y = total_peak_length)) +
  geom_point()


?geom_point
# here we are just looking at a x, y point plot (geom_point).
# note there are not quotes on x, and y objects as they are being evaluated as "objects in ggplot"
# note the +, we will see this everytime we are adding a new layer. # Yet within each layer it's standard comma 
# seperated etc.
# '+' adds layers together

             
# There is lots more we could modify here but let's change the    # shape of the points to show off R plotting. We need to do this  # outside the aes() layer. The aes is just for anything for       # that depends on any column of data. 

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
         
           geom_point(shape = 'square',
             color = 'purple')


# what other shapes are there and how would you find out ??????

# Now the points are plotted as squares.

# What if we wanted to color by TF or not TF column? Above, we "set" the 
# color and shape AFTER the aes(). To color by data that needs to be 
# evaluated we need to have that color INSIDE aes()

# DATA DEPENDENCE COLOR / not in GEOM layer
?aes
ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length
                )) +
  facet_wrap(tf ~ .) +
  geom_point()

# Now color came in aes() because it depends on data values.

# so much can be done in the Aesethics mapping layer!

# We can also add some color features in the mapping layer of graphical grammar let's take a look
# First let's see all the enteries
table(num_peaks_df$dbd)
# Ok there are several but most have very few enteries. Let's find those with at least
# 100 DBPs with a given DBD.

abundant_dbds <- table(num_peaks_df$dbd) > 100
names(which(abundant_dbds))
# we are checking which DBDs have more than 100 enteries or
# we have at least 100 DBPs with a given DBD.
# Ok we can see only the C2H2 Zinc Finger domain has more than 100 enteries.


# FUNCTION RUN in AES
      # num_peaks_df$dbd == "C2H2 ZF"
# here is the computation we are going to do in the aes layer
# we are going to see above how to take a list of characters
# and turn it into a boolean vector (T/F). Let's go over above
# and see how we can just add to the aes below.


ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length,
                 color = dbd == "C2H2 ZF")) +
  geom_point()


# We can see color outside aes() and we also see a function running!
# It's subtle but the == means that dbd will turn from a list of 
# characters to a vector boolean values (true or false). Let's look


# Finally note ggplot gives us a free title and legend!

```


# Histogram. This is great when you are inloadterested how often one variable occurs. So you would only use "x"
# or one data value in this case.

```{r}

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_histogram()

# let's add more bins

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_histogram(bins = 50)

# Let's chage the color of the bars in histogram by number of peaks.
# again we need to do this inside aes() since the data needs to read.

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
  geom_histogram(bins = 50)

# this is hard to see the two distributions.
# let's fix this by adding a 'position' parameter in the geom layer.

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_histogram(bins = 50, position = "dodge")

# so now we can see the individual data points seperated!

# DENSITY PLOTS # 
# DENISTY Plots are also very useful to see where the denisty of  # data arises.

# let's take a look at density plots comparing those that are and 
# are not TFs

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_density()

# but again its hard to see let's fix that with shading or alpha in
# geom layer

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf)) +
   geom_density(alpha = 0.1)

# let's change color of the line now (inside aes())

ggplot(num_peaks_df, aes(x = num_peaks, fill = tf, color = tf)) +
   geom_density(alpha = 0.3)

# so now line and fill are same color -- let's see:

ggplot(num_peaks_df, aes(x = num_peaks, color = tf)) +
   geom_density()

# We can even add multiple geom layers. For example a 2 dimensional denisty plot of our x, y point plot above.
# Let's check it out

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
  geom_point()+
  geom_density_2d() 

# what would happen if we change the order?

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +
   geom_density_2d()+
    geom_point()

# it's hard to see but in this case the contour is on top
# of points and points are below -- subtle but important to       # remember

```

You can do a lot in the geometry layer and add features with + and then new geom call. Let's look at 
# this below


```{r}
# Now let's see how to add other features to the plots by adding an additional geom layer as we did above
# Lets explore 'geom_abline' geometry to add a line 
?geom_abline
# we usually give slope and or intercept

ggplot(num_peaks_df, aes(x = num_peaks, 
                 y = total_peak_length)) +

  geom_point() +
  geom_abline(slope = 1000, intercept = 0) +
  geom_smooth(method = "gam")

?geom_smooth


# cool this fits -- what does it mean?
# hint average peaks size is 1kb

```

Let's take a quick look at bar plots
```{r}
# Here we will look at more aspects of graphical grammar with bar plots.
# Will plot all the types of DBD's so it's not overwhelming with 30,000+
# promoters etc.

ggplot(num_peaks_df, aes(x = dbd)) + 
  geom_bar()

  
# If you want the heights of the bars to represent values in the 
# data, use 'stat = identity' and provide that identity in the
# y = statement in aes.

# If the we want to provide the Y-axis for a specific number
# we need to apply stat = identitiy in the geom_bar()

ggplot(num_peaks_df, aes(
      x = dbp,
      y = num_peaks)) +
        geom_bar(stat = "identity")

```

Now we have gone through these layers:

1) Data -- data being plotted from object indexes etc.

2) MAPPING -- mapping data (aes)

3) Statistics -- statical analyses on mapped data

5) Geom

# Now let's look at the scales layer

4) scales -- how big small you want things
# Scales is important for outputting specific factors as colors that are "scaled"
# alos other outputs need to be scaled such as scale values to the "mean" etc.
?scale

```{r}

  
# Ok now let's see how we can use the scales layer to change axes.
#Let's look of using the scales layer by sclaing the axes

ggplot(num_peaks_df, aes(x = log10(num_peaks), y =log10(total_peak_length), color = tf)) +
  geom_point()+
  scale_x_continuous(breaks = c(1e3, 1e4, 1e5, 1e6)) +
  scale_color_brewer(palette = 3) 


#? scale_x_continious (same for y) is calling the scale layer specifically
# for the axes in ggplot. We are giving it where to put values on X-axis     # (note if a value is not met in largest break that break isn't printed)
# For the y-axis we are using transform (trans) to make it log 10 scaled


# what if we want to set the limits of the axes? The scales layer is
# just the place.

ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length, color = tf)) +
  geom_point()+
  xlim(250, 2e3) +

  ylim(5, 6) +
  scale_color_brewer(palette =  2) 


# so we add to scale functions with xlim and ylim (?xlim, ?ylim)

# coool so we zoomed in!

```

Now let's look at 6) FACETS
This will allow us to make multiple plots in the same plot frame or facet
One limitation is that it can only make the same plots from different data

FACET_GRID & FACET_WRAP
```{r}
# You have to be careful in the facet layer that you don't make too many
# plots. Essential we make as many figures as there are enteries.
# We will setd dbds to be the three categories of DBDs we want to facet.

# ?facet_grid

table(num_peaks_df$dbd)[order(table(num_peaks_df$dbd))]
# First let's take a look at the top three represented DBDs?
# Ok so now we will set dbds to these three DBDs.
dbds <- c("C2H2 ZF", "Homeodomain", "bZIP")

# Now the plotting begings and good example of running functions in aes()
ggplot(num_peaks_df %>% filter(dbd %in% dbds),
       # this is cool we can run function on df before 
       # pushing to aes()
       aes(x = num_peaks, y = total_peak_length )) +
  facet_grid(dbd ~ tf) + 
# here we call facet grid. The parameters are rows and columns seperated
# by ~. So we said take the the DBD and facet to see if it is a tf too.
# this will keep the x-axis the same and plot each DBD as yes or no if it
# is a tf as well. Almost all these domains are TFs so sparse no plots.
  geom_point()
# compare facet wrap

ggplot(num_peaks_df %>% filter(dbd %in% dbds), 
       aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(dbd ~ tf) + 
  geom_point()

# difference all combos of each factor. Same as before but facet wrap
# doesn't mind if all the values are in one category. So we see 
# " bZIP Yes", but not an equivelant NO factor since all bZIP doamins
# are TFs. Interestingly, one homeodomain has very few peaks and is not
# a TF.


# DEEPER INTO FACET #

# first let's just look at facet_wrap
ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ .) + 
# this notation . means facet_wrap the rows of tf but not 
# against another column as we did above. Just asking if a TF
# then it facets to the values that can be tabulated for TF
# which is just yes or no -- so we will get two plots. 
  
  geom_point()

# here we change the ordering of the figures. We currently have 1 
# row and two figures but we could easily make that two columns
# so the figures are stacked on top of each other. This is done
# in the facet layer inside facet_wrap as 'nrow' parameter. It
# defaults to one row until space is filled, then makes a new row.
ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ ., nrow = 2) + 
  geom_point()

# now let's have them scale themselves using scales = free.
# In facet grid one axis has to be shared, but in facet wrap we 
# can autoscale both axes. We can also tell it the number of 
# rows or columns we want. Let's see below.

ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length )) +
  facet_wrap(tf ~ ., scales = "free", ncol = 1) + 
  geom_point()
# note we are using ncol here istead of nrow you can use either one
# this would be nrow = 2 just to show ncol parameter.

```

7) Coordinates
How to represent X and Y coordinates in different coordinate systems.

```{r}

# R has a bunch of built in "cordinate" systems. Cartesian is the most common
# and what we have been using. But you can also plot data in "polar" 
# or other coordinate systems. We can just add it in the "Coordinates layer"

ggplot(num_peaks_df, aes(x = dbd)) +
  geom_bar() +
  coord_polar()

# the values eminate from the center and are a "bar_plot" from the center.
# we can easily see as we did above C2H2 ZF most represented in our data.

# We can play around in the coordinate space too. For example to tell what is # the "theta" or circular dimension. Essentially the bars will make arcs
# from the center acroding to their theta value (doesn't really matter).

ggplot(num_peaks_df) +
  geom_bar(aes(x = dbd)) +
  coord_polar(theta = 'y')

?

```

8) Theme
This is a great way to package all the features you want, colors,
backgrounds, grid lines etc. We tend to use paper white as a defualt theme.

We made a defualt theme that we can source at the begining of the
document call _setup.R. Once loaded a "paper white" theme will be available.

Let's look at example from 10_our_data_ranges where we called the theme.
Note they don't have to be in the order we are learning them in :)

```{r}
ggplot(num_peaks_df, aes(x = num_peaks)) + 
  geom_density(alpha = 0.2, color = "#424242", fill = "#424242") +
  theme_paperwhite() +
  xlab(expression("Number of peaks")) +
  ylab(expression("Density")) +
  ggtitle("Promoter binding events",
          subtitle = "mRNA and lncRNA genes")



# we can see here how titles can easily be made with ggtile which
# is in the "theme" layer of things.

# There are a number of built-in themes in addition to 
# the one we've provided. You can see them by typing theme_ 
# then autocomplete and check a few out! Here is cleavland :

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_density() +
  theme_cleveland()


# Go take a look at _setup.R and see the features of paper white!
```


