---
title: "05_bump_properties_template"
author: "Katy Walsh"
date: "4/21/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(ggplot2)
library(tidyverse)
source("../../util/plotting_functions.R")
source("../../util/_setup.R")
source('../../util/intersect_functions.R')
```


```{r importing}


```


Defining low binder and high binder promoters
```{r}

```

make TSS plots of histone marks for second hump 
Is there a corrolation between promoters with lots of DBPs bound vs those with fewer DBPs?
    Are there more DBPs binding because more chomatin is open? (or maybe do all the DBP binding events open or maintain open chromatin?)
```{r histone marks}



```


Are the genes belonging to the promoters with high DBP binding share common features?
    Do they have more exons compared to low bound promoters?
    Are they longer on average?
    Are there more mRNA than lncRNA or vise versa?
    differences in CpG islands? (would need to import data from UCSC)
```{r gene features}


```


Is there different levels of expression in the low bound promoters vs high bound promoters?
    Do reservoirs make up most of the second bump or are they distributed?
    (bring in RNA-seq data)
```{r expresion levels}

```

