---
title: "00_consensus_peaks"
author: "James Pratt"
date: "4/14/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(GenomicRanges)
source("/scratch/Shares/rinnclass/CLASS_2021/util/plotting_functions.R")
```

Note: pay attention to file paths


First, download data to desired directory ('/scratch/Shares/rinnclass/data/XXXX' in our case) using wget.

Website: https://www.dropbox.com/sh/76etjxl10v2hcuc/AAAIk9a6aMhw8CvUtKlFnMn6a?dl=0

Create and export consensus peaks from individual peak files.

```{r}
consensus_peaks <- create_consensus_peaks("/scratch/Shares/rinnclass/data/peaks")


for(i in 1:length(consensus_peaks)) {
  rtracklayer::export(consensus_peaks[[i]], 
                      paste0("/scratch/Shares/rinnclass/YOUR_FOLDER/X/results/chipseq/consensus_peaks/", 
                             names(consensus_peaks)[i], 
                             "_consensus_peaks.bed"))
}

```


Filtering the consensus peaks to files with at least 250 peaks (per last years class cut-off)
```{r}

num_peaks_threshold <- 250

filtered_consensus_peaks <- consensus_peaks[sapply(consensus_peaks, length) > num_peaks_threshold]

#Export consensus peaks
for(i in 1:length(filtered_consensus_peaks)) {
  rtracklayer::export(filtered_consensus_peaks[[i]], 
                      paste0("./results/chipseq/filtered_consensus/", 
                             names(filtered_consensus_peaks)[i], 
                             "_filtered_consensus_peaks.bed"))
}


```

Exporting Genomic features that we will use frequently and save to results folder

1) Promoter Regions
```{r}
gencode_gr <- rtracklayer::import("/Shares/rinn_class/data/genomes/human/gencode/v32/gencode.v32.annotation.gtf")

lncrna_mrna_promoters <- get_promoter_regions(gencode_gr, biotype = c("lncRNA", "protein_coding"))
rtracklayer::export(lncrna_mrna_promoters, "./results/lncrna_mrna_promoters.gtf")


lncrna_promoters <- get_promoter_regions(gencode_gr, biotype = "lncRNA")
rtracklayer::export(lncrna_promoters, "./results/lncrna_promoters.gtf")


mrna_promoters <- get_promoter_regions(gencode_gr, biotype = "protein_coding")
rtracklayer::export(mrna_promoters, "./results/mrna_promoters.gtf")
```


2) Gene Bodies

```{r}
lncrna_mrna_genebody <- gencode_gr[gencode_gr$type == "gene" & 
                                     gencode_gr$gene_type %in% c("lncRNA", "protein_coding")]
rtracklayer::export(lncrna_mrna_genebody, "./results/lncrna_mrna_genebody.gtf")


lncrna_genebody <- gencode_gr[gencode_gr$type == "gene" & 
                                gencode_gr$gene_type %in% c("lncRNA")]
rtracklayer::export(lncrna_genebody, "./results/lncrna_genebody.gtf")


mrna_genebody <- gencode_gr[gencode_gr$type == "gene" & 
                              gencode_gr$gene_type %in% c("protein_coding")]
rtracklayer::export(mrna_genebody, "./results/mrna_genebody.gtf")

```



