---
title: "02_designfile.Rmd"
author: "Group 2"
date: "10/15/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ENCODExplorer)
library(tidyverse)
library(janitor)
library(httr)
```

Michaels Encode Explorer

```{r}
contstruct_query <- function(experiment_accession,
                             base_url = "https://www.encodeproject.org/report.tsv?",
                             file_format = "fastq",
                             type = "File", 
                             status = "released",
                             fields = c("accession", "read_count", "md5sum",
                                        "controlled_by", "paired_end",
                                        "paired_with", "replicate", "target", "read_length")) {
  query <- paste(list(paste0("type=", type), 
        paste0("status=", status),
        paste0("file_format=", file_format),
        paste0("dataset=%2Fexperiments%2F", experiment_accession, "%2F"),
        map_chr(fields, ~paste0("field=", .))) %>%
          flatten(),
       collapse = "&")
  url <- paste0(base_url, query)
  return(url)
}
get_fastq_info <- function(experiment_accession) {
  request <- GET(contstruct_query(experiment_accession)) 
  body <- read_tsv(content(request, "text"), skip = 1)
  return(body)
}
contstruct_query("ENCSR516DDO")
resp <- GET("https://www.encodeproject.org/report.tsv?type=File&status=released&file_format=fastq&dataset=%2Fexperiments%2FENCSR516DDO%2F&field=accession&field=read_count&field=md5sum&field=controlled_by&field=paired_end&field=paired_with&field=replicate&field=target&field=read_length")
read_tsv(content(resp, "text"), skip = 1)
options(stringsAsFactors = FALSE)
# Read in the experiment report downloaded from Encode
# we will only use the Accession column 
# Here we'll retrieve the read_count and md5 sum 
# as well as the file accession (with some other metadata)
# for the fastq files associated with each experiment.
fastq_info <- read.table("data/class_encode_report.tsv",
                        skip = 1, sep = "\t", header = T) %>%
  dplyr::select(Accession) %>%
  rename(experiment_accession = Accession) %>%
  distinct() %>%
  mutate(fastq_info = map(experiment_accession, ~get_fastq_info(.))) %>%
  unnest(fastq_info) %>%
  rename(file_accession = Accession) %>%
  clean_names()
write_csv(fastq_info, "data/design_input.csv")
md5_txt <- fastq_info %>%
  mutate(fastq_file = paste0(file_accession, ".fastq.gz")) %>%
  dplyr::select(md5sum, fastq_file)
write.table(md5_txt, "data/fastq/md5.txt", sep = "  ", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

Awesome now we have a text file of encode file names and md5sum.
Let's run the md5sum test using -c to check a list and print the output to md5_status.txt

```
md5sum -c md5.txt 
md5sum md5sum -c md5.txt > md5_status.txt
```


TODO : Decide on fastQ merge as next step -- explore the table in excel to get a feel?

First we need to install "janitor"

Here we are going to use our ENCODE.TSV report to make a design file for 
Nextflow ChiP-seeker. You will need the report file that can be downloaded here:

https://www.encodeproject.org/report/?type=Experiment&status=released&assay_slims=DNA+binding&biosample_ontology.term_name=HepG2&assay_title=TF+ChIP-seq&biosample_ontology.classification=cell+line&files.read_length=100&files.read_length=76&files.read_length=75&files.read_length=36&assay_title=Control+ChIP-seq&assay_title=Histone+ChIP-seq&files.run_type=single-ended

Notice that the URL is very similar to the structure of the data file.

For the input design file Nextflow needs a few things:

group	replicate	fastq	antibody	control

* group is the DBP and controls listed individually
* replicate indicates which replicate of the given DBP 
* Antibody same for all DBPs but input controls are blank
* Control tells which "group" are the inputs.


Now we start the hard part of coding such a simple sounding design file :)


We are going to use the ENCODE API instead of the web interface. Some information
cannot be retrieved such as MD5sum and other aspects in the .TSV

You can see all the information you can gather using the API and what the code
would look like here:

https://app.swaggerhub.com/apis-docs/encodeproject/api/basic_search#/Experiment

Notice the URL produced above has a lot of the information -- so we can use the 
API to tell it exactly what we want in code rather than clicking buttons --


First let's make a table with the 
```{r}
# write_csv(design, "design_full_file")
## Grabbing some handy environment variables to .CSV :)
contstruct_query <- function(experiment_accession,
                             base_url = "https://www.encodeproject.org/report.tsv?",
                             file_format = "fastq",
                             type = "File", 
                             status = "released",
                             fields = c("accession", "read_count", "md5sum",
                                        "controlled_by", "paired_end",
                                        "paired_with", "replicate", "target")) {
  query <- paste(list(paste0("type=", type), 
        paste0("status=", status),
        paste0("file_format=", file_format),
        paste0("dataset=%2Fexperiments%2F", experiment_accession, "%2F"),
        map_chr(fields, ~paste0("field=", .))) %>%
          flatten(),
       collapse = "&")
  url <- paste0(base_url, query)
  return(url)
}
get_fastq_info <- function(experiment_accession) {
  request <- GET(contstruct_query(experiment_accession)) 
  body <- read_tsv(content(request, "text"), skip = 1)
  return(body)
}
# Read in the experiment report downloaded from Encode
# we will only use the Accession column 
# Here we'll retrieve the read_count and md5 sum 
# as well as the file accession (with some other metadata)
# for the fastq files associated with each experiment.
unique(missing_control_file)
# We can do another encode query to get the experiment control_type
# if we query experiments instead of files.
experiment_report <- read.table("data/Best_class_full_controls.tsv",
                        skip = 1, sep = "\t", header = T)
fastq_info <- data.frame(experiment_accession = c(experiment_report$Accession, c("ENCSR000EVT",
                                                                                 "ENCSR945JOR",
                                                                                 "ENCSR476TKW",
                                                                                 "ENCSR770PLF"))) %>%
  distinct() %>%
  mutate(fastq_info = map(experiment_accession, ~get_fastq_info(.))) %>%
  unnest(fastq_info) %>%
  dplyr::rename(file_accession = Accession) %>%
  clean_names()
write_csv(fastq_info, "final_files_md5sum.csv")
fastq_info <- read_csv("final_files_md5sum.csv")
```



Now that we have all the information we want in tables above we can tidy up
file names with gsub. So the first part is just cleaning the naming we made above
?gsub



# TODO # I think this is the new way of making the batch design files. So together
with chunk above maybe all we need.

```{r}
fastq_info$fastq_1 <- paste0("../fastq/", fastq_info$file_accession, ".fastq.gz")
fastq_info$file_control <- gsub("/", "", gsub("/files/", "", fastq_info$controlled_by))
fastq_info$fastq_2 <- ""
fastq_info$antibody <- gsub("-human", "", gsub("/", "", gsub("/targets/", "", fastq_info$target)))
fastq_info$group <- fastq_info$antibody
contrls <- fastq_info %>% filter(is.na(file_control))
contrls$group <- contrls$experiment_accession
control2exp <- fastq_info %>% 
  ungroup() %>% 
  dplyr::select(file_control) %>%
  distinct() %>%
  filter(!is.na(file_control)) %>%
  separate(file_control, into = c("one_control"), sep = ",", remove = FALSE)
ctrl2exp <- contrls %>%
ungroup() %>% 
  dplyr::select(experiment_accession, file_accession) %>%
  dplyr::rename(one_control = file_accession)
control2exp <- merge(control2exp, ctrl2exp)
control2exp <- control2exp %>%
  dplyr::select(-one_control) %>%
  dplyr::rename(control = experiment_accession)
fastq_info <- merge(fastq_info, control2exp, all.x = T)
fastq_info[is.na(fastq_info$controlled_by), "group"] <- fastq_info[is.na(fastq_info$controlled_by), "experiment_accession"]
fastq_info <- fastq_info %>% 
  group_by(group) %>%
  mutate(replicate = 1:n())
design <- fastq_info %>%
  dplyr::select(group, replicate, fastq_1, fastq_2, antibody, control)
experiments <- design %>% 
  filter(!is.na(control)) %>%
  dplyr::select(group) %>%
  distinct() 
experiments$exp_num <- 1:nrow(experiments)
experiments$batch <- as.numeric(cut_number(experiments$exp_num, n = 6))
experiments <- experiments %>% dplyr::select(group, batch) %>%
  as.data.frame()
design <- merge(design, experiments, all.x = T)
batches <- design %>%
  group_by(batch) %>%
  group_split()
controls <- batches[[7]]
for (i in 1:6) {
  batch <- batches[[i]]
  ctrl_batch <- controls[controls$group %in% batch$control,]
  batch <- bind_rows(ctrl_batch, batch)
  # Change the NAs to ""
  batch[is.na(batch$antibody),"antibody"] <- ""
  batch[is.na(batch$control),"control"] <- ""
  # Get rid of batch col
  batch <- batch %>% dplyr::select(-batch)
  dir.create(paste0("data/test_work/batch_",i))
  write_csv(batch, paste0("data/test_work/batch_",i,"/design_",i,".csv"))
}
```