---
title: "Create consensus peaks"
output: html_document
editor_options: 
  chunk_output_type: console
---
KW version
group: James, Victor, Nina

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(GenomicRanges)
source("/scratch/Shares/rinnclass/Katy/CLASS_2021/util/intersect_functions.R")
```

creating and exporting consensus peaks from individual peak files
place these files in "CLASS_2021/analysis/00_consensus_peaks/results/" directory path

The key here is to place all the data we will use moving forward in your directory 

May your attention to file paths serve you well !
```{r}
consensus_peaks <- create_consensus_peaks("/scratch/Shares/rinnclass/data/peaks")


for(i in 1:length(consensus_peaks)) {
  rtracklayer::export(consensus_peaks[[i]], 
                      paste0("/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/consensus_peaks/", 
                             names(consensus_peaks)[i], 
                             "_consensus_peaks.bed"))
}

```


Filtering the consensus peaks to files with at least 250 peaks (per last years class cut-off)
```{r}

num_peaks_threshold <- 250

filtered_consensus_peaks <- consensus_peaks[sapply(consensus_peaks, length) > num_peaks_threshold]

#Export consensus peaks
for(i in 1:length(filtered_consensus_peaks)) {
  rtracklayer::export(filtered_consensus_peaks[[i]], 
                      paste0("/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/filtered_consensus/", 
                             names(filtered_consensus_peaks)[i], 
                             "_filtered_consensus_peaks.bed"))
}


```


Exporting Genomic features that we will use frequentyl so we can use rtracklayer::import and save a ton of time :)

These you will want to save directly to your results folder since they will be used for all analyses and not just chipseq (where we exported consensus and filtered consensus peaks)

1) Promoter Regions
```{r}
gencode_gr <- rtracklayer::import("/Shares/rinn_class/data/genomes/human/gencode/v32/gencode.v32.annotation.gtf")

lncrna_mrna_promoters <- get_promoter_regions(gencode_gr, biotype = c("lncRNA", "protein_coding"))
rtracklayer::export(lncrna_mrna_promoters, "/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/gene_annotation/lncrna_mrna_promoters.gtf")


lncrna_promoters <- get_promoter_regions(gencode_gr, biotype = "lncRNA")
rtracklayer::export(lncrna_promoters, "/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/gene_annotation/lncrna_promoters.gtf")


mrna_promoters <- get_promoter_regions(gencode_gr, biotype = "protein_coding")
rtracklayer::export(mrna_promoters, "/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/gene_annotation/mrna_promoters.gtf")
```


2) Gene Bodies

```{r}
lncrna_mrna_genebody <- gencode_gr[gencode_gr$type == "gene" & 
                                     gencode_gr$gene_type %in% c("lncRNA", "protein_coding")]
rtracklayer::export(lncrna_mrna_genebody, "/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/lncrna_mrna_genebody.gtf")


lncrna_genebody <- gencode_gr[gencode_gr$type == "gene" & 
                                gencode_gr$gene_type %in% c("lncRNA")]
rtracklayer::export(lncrna_genebody, "/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/lncrna_genebody.gtf")


mrna_genebody <- gencode_gr[gencode_gr$type == "gene" & 
                              gencode_gr$gene_type %in% c("protein_coding")]
rtracklayer::export(mrna_genebody, "/scratch/Shares/rinnclass/Katy/CLASS_2021/analysis/00_consensus_peaks/results/mrna_genebody.gtf")

```



